---
title: "Compare GTSM and wind"
author: "Willem Stolte, Nathalie Dees"
date: "2024-09-12"
params:
  monitoryear: 2024
output:
  html_document:
    df_print: paged
    code_folding: hide
---

# Compare gtsm with wind

```{r setupWind}
knitr::opts_chunk$set(
  echo = TRUE,
  comment=FALSE,
  message=FALSE,
  warning = FALSE
)
```

```{r dependencies}
require(tidyverse)
source("../sealevelmonitor/_common/functions.R")
source("../sealevelmonitor/_common/plotfunctions.R")
config <- RcppTOML::parseToml("../sealevelmonitor/_common/configuration.TOML")
config_flat <- list_flatten(config)

```

```{r readNewGTSM}

gtsm <- read_yearly_gtsm(filename = "../../data/deltares/gtsm/gtsm_surge_annual_mean_main_stations.csv") |>
  mutate(
    year = year(ymd(t)),
    surge_mm = surge *1000
    )

```


## measured wind at De Kooy

KNMI daily average wind observations were downloaded via https://www.knmi.nl/nederland-nu/klimatologie/daggegevens.  . Data for station "De Kooy" ([station nr 235](https://cdn.knmi.nl/knmi/map/page/klimatologie/gegevens/daggegevens/etmgeg_235.zip)) were selected. This station is close by the tidal station "Den Helder". 


```{r readKNMIwind}
list.files("../../data/knmi/metingen")

# daggemiddelden (alleen gebruikt voor het berekenen van jaargemiddelden)
kooy_raw = read_delim("../../data/knmi/metingen/dekooy_wind.txt", comment = "#", trim_ws = T)

# uurgemiddelden voor gebruik in windrozen
kooy_uur_raw = read_delim("../../data/knmi/metingen/dekooy_uurgeg_wind.txt", comment = "#", trim_ws = T) %>%
  rename(
    DDVEC = DD,
    FHVEC = FH
  )

kooy = kooy_raw %>%
  mutate(
    year = YYYYMMDD %/% 10000,
    windspeed = as.numeric(FHVEC) / 10,
    metrad = as.numeric(DDVEC) * pi / 180,
    u = (windspeed) * (-sin(metrad)),
    v = (windspeed) * (-cos(metrad)),
       # Mathematical metrad
   rad = 2*pi - metrad - pi/2,
   rad = ifelse(rad >= 2*pi, rad - 2*pi, rad)
         ) %>%
  filter(year < params$monitoryear) %>%
  group_by(year, STN) %>%
  summarise(
    windspeed = mean(windspeed),
    u = mean(u),
    v = mean(v),
    rad = mean(rad),
    u2 = u^2 * sign(u),
    v2 = v^2 * sign(v)
            )

```




## Windrose

Windroses for the last four years show that there is considerable variation in average wind speed, and direction distribution.

```{r windrose,  fig.show="hold", out.width="50%", fig.cap="Windroses of measured wind at De Kooy station for the last 4 years. "}

yearsToPlot <-  seq(params$monitoryear - 4, params$monitoryear-1)

kooy_raw_2020_2023 = kooy_uur_raw %>% 
  mutate(
    year = YYYYMMDD %/% 10000,
    spd = as.numeric(FHVEC) / 10,
    dir = as.numeric(DDVEC)
  ) %>%
  filter(year %in% yearsToPlot) %>%
  select(year, spd, dir) %>%
  group_by(year) %>%
  nest()

walk(1:length(kooy_raw_2020_2023$data), 
     function(x) {
       plot.windrose(
         spd = kooy_raw_2020_2023$data[[x]]$spd, 
         dir = kooy_raw_2020_2023$data[[x]]$dir,
       plot.title = kooy_raw_2020_2023$year[x],
       palette = "YlOrBr"
       )
     }
)

```

### ERA5 reanalysis 


```{r readERA5wind}


eraURL <- "data/copernicus/era5/era5_wind_for_analysis.csv"

era5wind <- read_delim(file.path("../../", eraURL)) %>%
  rename(
    u = u_mean,
    v = v_mean,
  ) %>%
  mutate(
    year = year(year),
    u2 = u^2 * sign(u),
    v2 = v^2 * sign(v),
    source = "ERA5"
    ) %>%
  filter(year < params$monitoryear)

```


```{r compareERA5-KNMI}

era5wind %>%
  filter(stationname == "DEN HELDER") %>%
  bind_rows(kooy %>% mutate(stationname = "DEN HELDER") %>% mutate(source = "KNMI")) %>%
  pivot_longer(cols = c(u,v), names_to = "component", values_to = "windspeed_ms") %>%
  ggplot(aes(year, windspeed_ms)) +
  geom_line(aes(color = source), linewidth = 1) +
  facet_wrap("component", ncol = 1) +
  ggtitle(
    label = "Annual average wind",
    subtitle = "ERA5 reanalysis for Den Helder - KNMI measured wind at De Kooy."
    ) +
  theme_minimal()
  # left_join(kooy, by = c(year = "year")) %>%
  

```




## Wind components

Wind speed and direction were parsed in two components *u* and *v* perpendicular to eachother. *u* represents the component from west to east, *v* the component from south to north. Furthermore, the quadratic wind speeds *u^2* and *v^2* were compared with surge, since wind force is proportional to the square of wind speed. 


### Compare GTSM with KNMI measured wind at Den Helder 

```{r compareGTSM_KNMI, figures-side, fig.show="hold", out.width="50%"}

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(kooy) %>%
  ggplot() +
  geom_point(aes(u2, surge_mm)) +
  geom_text(aes(u2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 1) +
  geom_smooth(aes(u2, surge_mm, label = year), method = 'lm')

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(kooy) %>%
  ggplot() +
  geom_point(aes(v2, surge_mm)) +
  geom_text(aes(v2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 1) +
  geom_smooth(aes(v2, surge_mm, label = year), method = 'lm')

```



```{r}

gtsm %>%
  filter(name == "Den Helder") %>%
  mutate(surge_cm = surge_mm / 10) %>%
  left_join(kooy) %>%
  filter(year < 2024) %>%
  select(year, u2, v2, surge_cm) %>%
  pivot_longer(c(u2, v2, surge_cm), names_to = "factor", values_to = "value") %>%
  ggplot() +
  geom_path(aes(year, value, color = factor), size = 1)


```

### Compare GTSM with ERA5 reanalysis wind at Den Helder 


```{r compareGTSM_ERA5, figures-side, fig.show="hold", out.width="50%"}

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year") ) %>%
  ggplot() +
  geom_point(aes(u2, surge_mm)) +
  geom_text(aes(u2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 0) +
  geom_smooth(aes(u2, surge_mm, label = year), method = 'lm') +
  ggtitle(label = NULL, subtitle = "GTSM surge and ERA5 wind reanalysis component u")

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year") ) %>%
  ggplot() +
  geom_point(aes(v2, surge_mm)) +
  geom_text(aes(v2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 0) +
  geom_smooth(aes(v2, surge_mm, label = year), method = 'lm') +
  ggtitle(label = NULL, subtitle = "GTSM surge and ERA5 wind reanalysis component v")

```

```{r}

gtsm %>%
  filter(name == "Den Helder") %>%
  mutate(surge_cm = surge_mm / 10) %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year")) %>%
  filter(year < 2024) %>%
  select(year, u2, v2, surge_cm) %>%
  pivot_longer(c(u2, v2, surge_cm), names_to = "factor", values_to = "value") %>%
  ggplot() +
  geom_path(aes(year, value, color = factor), size = 1)


```



