---
date: "2024-09-12"
params:
  monitoryear: 2025
output:
  github_document:
  html_document:
    df_print: paged
    code_folding: hide
always_allow_html: true
---

# Compare GTSM with wind measurements and reanalyses

```{r setupWind, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment=FALSE,
  message=FALSE,
  warning = FALSE
)

require(tidyverse)
require(DT)
source("../sealevelmonitor/_common/functions.R")
source("../sealevelmonitor/_common/plotfunctions.R")
config <- RcppTOML::parseToml("../sealevelmonitor/_common/configuration.TOML")
config_flat <- list_flatten(config)

```


## Global Tide and Surge Model

The GTSM model results for wind surge were produced by Deltares (Muis et al., 2016). Wind surge is extracted for the segement representative for each one of the Dutch main tide gauge stations, and averaged per year. This results in wind surge per year and station. Based on previous comparisons with alternative corrections for wind, GTSM was chosen as the preferential method (Deltares 2023). 

A first glance at the GTSM wind surge results is shown below. 

```{r readNewGTSM}

gtsm <- read_yearly_gtsm(filename = "../../data/deltares/gtsm/gtsm_surge_annual_mean_main_stations.csv") |>
  mutate(
    year = year(ymd(t)),
    surge_mm = surge *1000
    ) %>% 
  select(-t, -surge)

kableExtra::kable(align = "c",
    gtsm %>% arrange(-year),
    caption = "GTSM wind surge per station and year.",
    digits = 2
    ) %>%
  kableExtra::scroll_box(height = "500px") %>%
  kableExtra::kable_styling(full_width = T)

```

For validation of GTSM wind surge, and perhaps a first explanation of its variation, GTSM surge is compared to measured wind along the Dutch coast, and a commonly used reanalysis of wind selection for the Dutch coast. 

## measured wind at De Kooy

KNMI daily average wind observations were downloaded via https://www.knmi.nl/nederland-nu/klimatologie/daggegevens.  . Data for station "De Kooy" ([station nr 235](https://cdn.knmi.nl/knmi/map/page/klimatologie/gegevens/daggegevens/etmgeg_235.zip)) were selected. This station is close by the tidal station "Den Helder". Wind data, originally as wind direction and wind speed, were decomposed in its two perpendicular components ***u*** and ***v***. 


```{r readKNMIwind}
list.files("../../data/knmi/metingen")

# daggemiddelden (alleen gebruikt voor het berekenen van jaargemiddelden)
kooy_raw = read_delim("../../data/knmi/metingen/etmgeg_235.txt", comment = "#", trim_ws = T)

# uurgemiddelden voor gebruik in windrozen
kooy_uur_raw = read_delim("../../data/knmi/metingen/dekooy_uurgeg_wind.txt", comment = "#", trim_ws = T) %>%
  rename(
    DDVEC = DD,
    FHVEC = FH
  )

kooy = kooy_raw %>%
  mutate(
    year = YYYYMMDD %/% 10000,
    windspeed = as.numeric(FHVEC) / 10,
    metrad = as.numeric(DDVEC) * pi / 180,
    u = (windspeed) * (-sin(metrad)),
    v = (windspeed) * (-cos(metrad)),
       # Mathematical metrad
   rad = 2*pi - metrad - pi/2,
   rad = ifelse(rad >= 2*pi, rad - 2*pi, rad)
         ) %>%
  filter(year < params$monitoryear) %>%
  group_by(year, STN) %>%
  summarise(
    windspeed = mean(windspeed),
    u = mean(u),
    v = mean(v),
    rad = mean(rad),
    u2 = u^2 * sign(u),
    v2 = v^2 * sign(v)
            )

```




## Windrose

Windroses for the last four years show that there is considerable variation in average wind speed, and direction distribution.

```{r windrose,  fig.show="hold", out.width="50%", fig.cap="Windroses of measured wind at De Kooy station for the last 4 years. "}

yearsToPlot <-  seq(params$monitoryear - 4, params$monitoryear-1)

kooy_raw_last_4_years = kooy_uur_raw %>% 
  mutate(
    year = YYYYMMDD %/% 10000,
    spd = as.numeric(FHVEC) / 10,
    dir = as.numeric(DDVEC)
  ) %>%
  filter(year %in% yearsToPlot) %>%
  select(year, spd, dir) %>%
  group_by(year) %>%
  nest()

walk(1:length(kooy_raw_last_4_years$data), 
     function(x) {
       plot.windrose(
         spd = kooy_raw_last_4_years$data[[x]]$spd, 
         dir = kooy_raw_last_4_years$data[[x]]$dir,
       plot.title = kooy_raw_last_4_years$year[x], 
       palette = "YlOrBr"
       )
     }
)

```

### ERA5 reanalysis 


```{r readERA5wind}


eraURL <- "data/copernicus/era5/era5_wind_for_analysis.csv"

era5wind <- read_delim(file.path("../../", eraURL)) %>%
  rename(
    u = u_mean,
    v = v_mean,
  ) %>%
  mutate(
    year = year(year),
    u2 = u^2 * sign(u),
    v2 = v^2 * sign(v),
    source = "ERA5"
    ) %>%
  filter(year < params$monitoryear)

```


```{r compareERA5-KNMI}

era5wind %>%
  filter(stationname == "DEN HELDER") %>%
  bind_rows(kooy %>% mutate(stationname = "DEN HELDER") %>% mutate(source = "KNMI")) %>%
  pivot_longer(cols = c(u,v), names_to = "component", values_to = "windspeed_ms") %>%
  ggplot(aes(year, windspeed_ms)) +
  geom_line(aes(color = source), linewidth = 1) +
  facet_wrap("component", ncol = 1) +
  ggtitle(
    label = "Annual average wind",
    subtitle = "ERA5 reanalysis for Den Helder - KNMI measured wind at De Kooy."
    ) +
  theme_minimal()
  # left_join(kooy, by = c(year = "year")) %>%
  

```




## Wind components

Wind speed and direction were parsed in two components *u* and *v* perpendicular to eachother. *u* represents the component from west to east, *v* the component from south to north. Furthermore, the quadratic wind speeds *u^2* and *v^2* were compared with surge, since wind force is proportional to the square of wind speed. 


### Compare GTSM with KNMI measured wind at Den Helder 

```{r compareGTSM_KNMI, figures-side, fig.show="hold", out.width="50%"}

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(kooy) %>%
  ggplot() +
  geom_point(aes(u2, surge_mm)) +
  geom_text(aes(u2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 1) +
  geom_smooth(aes(u2, surge_mm, label = year), method = 'lm')

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(kooy) %>%
  ggplot() +
  geom_point(aes(v2, surge_mm)) +
  geom_text(aes(v2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 1) +
  geom_smooth(aes(v2, surge_mm, label = year), method = 'lm')

```



```{r}

gtsm %>%
  filter(name == "Den Helder") %>%
  mutate(surge_cm = surge_mm / 10) %>%
  left_join(kooy) %>%
  filter(year < params$monitoryear) %>%
  select(year, u2, v2, surge_cm) %>%
  pivot_longer(c(u2, v2, surge_cm), names_to = "factor", values_to = "value") %>%
  ggplot() +
  geom_path(aes(year, value, color = factor), size = 1)


```

### Compare GTSM with ERA5 reanalysis wind at Den Helder 


```{r compareGTSM_ERA5, figures-side, fig.show="hold", out.width="50%"}

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year") ) %>%
  ggplot() +
  geom_point(aes(u2, surge_mm)) +
  geom_text(aes(u2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 0) +
  geom_smooth(aes(u2, surge_mm, label = year), method = 'lm') +
  ggtitle(label = NULL, subtitle = "GTSM surge and ERA5 wind reanalysis component u")

gtsm %>%
  filter(name == "Den Helder") %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year") ) %>%
  ggplot() +
  geom_point(aes(v2, surge_mm)) +
  geom_text(aes(v2, surge_mm, label = year), alpha = 0.5, nudge_x = 0.1, nudge_y = 0) +
  geom_smooth(aes(v2, surge_mm, label = year), method = 'lm') +
  ggtitle(label = NULL, subtitle = "GTSM surge and ERA5 wind reanalysis component v")

```

```{r}

gtsm %>%
  filter(name == "Den Helder") %>%
  mutate(surge_cm = surge_mm / 10) %>%
  left_join(era5wind %>% filter(stationname == "DEN HELDER"), by = c(year = "year")) %>%
  filter(year < 2024) %>%
  select(year, u2, v2, surge_cm) %>%
  pivot_longer(c(u2, v2, surge_cm), names_to = "factor", values_to = "value") %>%
  ggplot() +
  geom_path(aes(year, value, color = factor), size = 1)


```


# Literatuur

Muis, Sanne, Martin Verlaan, Hessel C. Winsemius, and Philip J. Ward Jeroen C. J. H. Aerts. 2016. “A Global Reanalysis of Storm Surges and Extreme Sea Levels.” Nature Communications, Articles 7. https://doi.org/10.1038/ncomms11969.
