---
title: "Untitled"
author: "Willem Stolte"
date: "2025-06-16"
output: github_document
---

## Introduction

One of the models used in the Sea Level Monitor is a broken linear model. In previous reports, the breakpoint was defined in the year 1993. This was done for two reasons:

- Analyses showed that global and Dutch sea level in the beginning of the 1990's was lower than expected. One of the explanations was a global temperature drop due to the Pinatubo eruption
- By choosing the year 1993, the trend of the part after the break could be easily compared to trends calculated from satellite observations, as they also start in this year. At the same time, the connection with earlier years is kept intact. 

This notebook explores the most likely breakpoint year. A strong deviation from 1993 would indicate that the breakpoint year may have to change, or that the model is no longer valid. 


```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  comment=FALSE,
  message=FALSE,
  warning = FALSE
)

require(leaflet)
require(plotly)
require(nlme)
require(rsealevel)

config <- RcppTOML::parseToml("_common/configuration.TOML")
# source("_common/functions.R")
source("_common/plotfunctions.R")
epoch = config$constants$epoch
mainstations_df <- readMainStationInfo(filepath = "../../")
mainstations_locs <- readMainStationLocations(path = "../../")


```

## Read sea level data

```{r readData, comment=F, message=F}

  df <- read_delim(file.path("../", "../data/deltares/input/psmsl_gtsm_yr-latest.csv"), delim = ";") %>%
  # filter(station %in% params$station) %>%
  filter(year >= 1890)

```

## Find breakpoints for broken linear regression

Note: no correction is made for nodal tide in this implementation.

```{r}
library(segmented)
```

Breakpoint function

```{r}

slm_breakpoint <- function(
    df, 
    mystation = "Netherlands (without Delfzijl)", 
    startyear = 1890, 
    endyear = 2025,
    n_bp, 
    printsummary = F
    ) {

  nr_of_breakpoints = 2
  
  mydf = df %>% 
    filter(
      station %in% mystation,
      year >= startyear,
      year <= endyear
    ) %>%
    mutate(corr_height = height - surge_anomaly) %>%
    group_by(year) %>%
    summarize(corr_height = mean(corr_height))
  
  fit_lm = lm(corr_height ~ year, data = mydf )  # intercept-only model
  
  fit_segmented = selgmented(
    fit_lm, 
    seg.Z = ~year,
    Kmax = nr_of_breakpoints,
    type = "aic",
    bonferroni = TRUE,
    plot.ic = TRUE
  )  # Two change points along x
  
 
  
  plot(mydf$year, mydf$corr_height, main = mystation)
  # lines(mydf$year, predict((fit_segmented)), add = T)
  
  plot(fit_segmented, add = T)
  # points(mydf)
  lines.segmented(fit_segmented)
  points.segmented(fit_segmented)
  
  if(printsummary){
    summary(fit_segmented)
  }
}
```


```{r runBreakpoint}

slm_breakpoint(df, 
               mystation = "Netherlands (without Delfzijl)",
               startyear = 1890,
               endyear = 2025,
               n_bp = 4, 
               printsummary = T
               )

```

From this analysis, the best fit is using one breakpoint at 1998 +/- 5 years. The currently used breakpoint (1993) is just within this standard error of the estimate. 


## surge 

Wind surge is determined with the Global Tide and Surge Model (GTSM). The following part tests whether ther is any breakpoint in the surge data. 


```{r}

surge_breakpoint <- function(
    df, 
    mystation = "Netherlands (without Delfzijl)", 
    startyear = 1890, 
    n_bp, 
    printsummary = T
    ) {

  nr_of_breakpoints = 2
  
  mydf = df %>% 
    filter(
      station %in% mystation,
      year >= startyear
    ) #%>%

  fit_lm = lm(surge_anomaly ~ year, data = mydf )  # intercept-only model
  
  fit_segmented = selgmented(
    fit_lm, 
    seg.Z = ~year,
    Kmax = nr_of_breakpoints,
    type = "aic",
    bonferroni = TRUE,
    plot.ic = TRUE
  )  # Two change points along x
  
  # plot(mydf$year, mydf$surge_anomaly, main = mystation)
  # # lines(mydf$year, predict((fit_segmented)), add = T)
  # 
  # plot(fit_segmented, add = T)
  # # points(mydf)
  # lines.segmented(fit_segmented)
  # points.segmented(fit_segmented)
  
  if(printsummary){
    summary(fit_segmented)
  }
}

```



```{r}

surge_breakpoint(df %>% filter(year >= 1950), 
               mystation = "Netherlands (without Delfzijl)",
               startyear = 1890,
               n_bp = 1, 
               printsummary = T
               )

```


There are no breakpoints detected in the surge anomaly time series. 

Only years from which surge data are available ( >= 1950)
