---
title: "GAM model estimation of Dutch Sea Level"
author: "Willem Stolte"
date: '`r Sys.Date()`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = FALSE
  )
source("_common/functions.R")
```

## Introduction

Full documentation will follow

The Dutch Sea Level Monitor uses general linearized model variants to describe the sea level at the Dutch coastal tidal stations. In the literature, also other models are used. For example, GAM was used by Keizer et al., (2023) to detect changes in sea level rise by comparing different periods with respect to GAM fitted sea level changes. Although GAM is at the moment not included in the official product, in this document we test this method in order to visualize patterns of sea level (rise). 

This notebook provides the code and first results


```{r loadInformation}
# load necessary library
library(mgcv)    # for GAMs
library(tidyverse)
library(gratia)

epoch = 1970
datapath = "../../data/deltares/input/psmsl_gtsm_yr-latest.csv"

df <- read_delim(file.path(datapath), delim = ";", col_types = cols()) %>%
  # filter(station %in% params$station) %>%
  filter(year >= 1890)

df <- df %>% 
  mutate(
    station = factor(
      station, levels = c(
        "Delfzijl",
        "Harlingen",
        "Den Helder",
        "IJmuiden",
        "Hoek van Holland",
        "Vlissingen",
        "Netherlands",
        "Netherlands (without Delfzijl)"
      )
    )
  )
```

## Define GAM 

The GAM model used here is using 4 components:

- smooth term
- nodal cosine term
- nodal sine term
- surge anomaly term

No further fine tuning of the GAM model has been done so far

## Add helper variables

```{r}

prepare_nodal <- function(df, epoch = 1970){
  df %>%
    mutate(
      nodal_cos = cos(2 * pi * (year - epoch) / 18.613),
      nodal_sin = sin(2 * pi * (year - epoch) / 18.613)
    )
}

```


## compare gam and glm

The gam and glm models can be compared when using the same model formulations. For this comparison, the linear model is also formulated using the mgcv package (like the gam model) but without the smoothing term, so only using linear terms. 

For that reason the exact outcomes of the GLM model in this notebook does not necessarily match exactly with the GLM outcomes in the Sea Level Monitor script. 



```{r}

slm_k = 50

gam_model <- function(df, epoch = 1970, k = slm_k){

  mgcv::gam(
    height ~ s(year, k = k)   # smooth term on year
    +     nodal_cos
    +     nodal_sin
    + surge_anomaly,
    family = gaussian(),
    data = df,
    method = "REML"
  )
}


glm_model <- function(df, epoch = 1970){

  mgcv::gam(
    height ~ year
    +        from1993 
    +        nodal_cos
    +        nodal_sin
    +        surge_anomaly,
    family = gaussian(),
    data = df,
    method = "REML"
  )
}

get_year_predictions <- function(model, newdata, modeltype){

  # get all term contributions + SE
  p <- predict(model, newdata = newdata, type = "terms", se.fit = TRUE)

  # GAM model → use only s(year)
  if (modeltype == "gam") {
    term_name <- "s(year)"
    fit  <- p$fit[, term_name]
    se   <- p$se.fit[, term_name]
  }

  # GLM model → use year + from1993
  if (modeltype == "glm") {
    terms_needed <- c("year", "from1993")

    fit  <- rowSums(p$fit[, terms_needed, drop = FALSE])

    # combine SE correctly (assuming independence of linear terms)
    se   <- sqrt(rowSums(p$se.fit[, terms_needed, drop = FALSE]^2))
  }

  tibble(
    fit_year = fit,
    se_year  = se
  )
}


```

## Apply GAM and GLM models to sea level data for all stations

The GAM and GLM models was applied to all six main stations and the composite stations according to the code below. 

```{r get-derivatives}

get_derivatives <- function(model) {

  # ==== CASE 1: GAM MET SMOOTH ====
  if (length(model$smooth) > 0) {

    der <- gratia::derivatives(
      model,
      term = "s(year)",
      type = "central",
      interval = "confidence",
      unconditional = TRUE
    )


  # Bouw een nette, uniforme tibble met de juiste kolommen
  out <- tibble::tibble(
    x          = der$year,
    derivative = der$.derivative,
    se         = der$.se,
    lower      = der$.lower_ci,
    upper      = der$.upper_ci
  )

    return(out)
  }

  # ==== CASE 2: LINEAIR MODEL ====

  cf <- coef(model)
  vc <- vcov(model)

  year_term <- grep("^year$", names(cf), value = TRUE)
  from_term <- grep("^from1993$", names(cf), value = TRUE)

  if (length(year_term) == 0) {
    stop("GLM heeft geen year-term (aliased in dit station).")
  }

  slope_year  <- cf[[year_term]]
  se_year     <- sqrt(vc[year_term, year_term])

  if (length(from_term) == 1) {
    slope_from <- cf[[from_term]]
    se_from    <- sqrt(vc[from_term, from_term])
  } else {
    slope_from <- 0
    se_from    <- 0
  }

  xvals <- model$model$year

  # ✅ derivative vóór/na 1993
  derivative_vals <- ifelse(
    xvals < 1993,
    slope_year,
    slope_year + slope_from
  )

  # SE combineren
  se_vals <- sqrt(se_year^2 + se_from^2)

  tibble::tibble(
    x          = xvals,
    derivative = derivative_vals,
    se         = se_vals,
    lower      = derivative_vals - 2*se_vals,
    upper      = derivative_vals + 2*se_vals
  )
}

```


```{r}

selected_model = c("glm", "gam")

by_station_model_compared = df %>%
  prepare_nodal() %>%
  addBreakPoints() %>%
  group_by(station) %>%
  nest() %>%
  ungroup() %>%
  expand_grid(modeltype = selected_model) %>% # in case multiple models are used
  #### parse function names based on selected model()
  mutate(modelfunctionname = paste(modeltype, "model", sep = "_")) %>%
  #### add functions for model calculation
  mutate(modelfunctions = map(modelfunctionname, get)) %>%
  #### add models based on data and functions
  mutate(model = pmap(
    list(
      data,
      modelfunctions
    ),
    \(.d, .f) .f(.d)
  )) %>%
  #### add output parameters and predictions
  mutate(
    glance = map(model, broom::glance),
    # rsq    = glance %>% map_dbl("r.squared"),
    adj.rsq = glance %>% map_dbl("adj.r.squared"),
    AIC    = glance %>% map_dbl("AIC"),
    npar = glance %>% map_dbl("npar"),
    tidy   = map(model, ~ broom::tidy(.x, parametric = TRUE, smooth = TRUE)),
    augment = map(model, broom::augment)#,
    # equation = map(model, function(x) equatiomatic::extract_eq(x))
  ) %>%
  mutate(
    pred_year = pmap(
      list(model, data, modeltype),
      ~ get_year_predictions(..1, ..2, ..3),
    ),
    derivatives = map(model, get_derivatives)
  ) 


```



## model comparison

There is something not correct with the offset of the models. They differ. This needs to be corrected later. For visualization, I project the models and data on top of eachother, but it needs some more attention as it also seems to affect the SE. 

```{r, fig.width=10, fig.height=8}
by_station_model_compared %>%
  select(data, pred_year, modeltype, station) %>%
  # filter(station == "Netherlands (without Delfzijl)") %>%
  unnest(c(data, pred_year)) %>% 
  group_by(modeltype, station) %>%
  mutate(
      pred_year_centered = fit_year - mean(fit_year),
      height_centered = height - mean(height)
  ) %>%
  ungroup() %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = height_centered, color = modeltype), alpha = 0.3) +
  geom_line(aes(y = pred_year_centered, color = modeltype), linewidth = 1) +
  # geom_ribbon(aes(ymin = pred_year_centered - se_year, ymax = pred_year_centered + se_year, fill = modeltype), alpha = 0.2) +
  facet_wrap("station")
```

## Sea level rate development in time

The comparison between observed and (GAM) modelled sea level for the 6 stations, and the composite stations is shown below. The variation in observations from 1950 to now is lower than in the period before. this is related to the wind-surge corrections derived from the GTSM model. Because of a lack of GTSM output for the years before 1950, for those years only a correction with the mean surge has been made. 

The plot below shows the sea level rate in time as estimated using GAM. It is currently compared to the sea level rate as calculated with the current Sea Level Monitor preferred model (broken linear) for the virtual station "Netherlands (without Delfzijl)". For better comparison on station level, station-specific results will be added in a later stage.  


```{r plot-derivatives, fig.cap="Derivatives plus confidence intervals.", fig.height=6, fig.width=8 }

plot_derivatives <- function(by_station_model_compared) {

  # unnest derivatives én metadata
  df_plot <- by_station_model_compared %>% 
    select(station, modeltype, derivatives) %>%
    unnest(derivatives)

  ggplot(df_plot, aes(x = x, y = derivative, color = modeltype)) +
    geom_line(size = 1) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = modeltype),
                alpha = 0.2, color = NA) +
    facet_wrap(~ station, scales = "free_y") +
    theme_bw() +
    labs(
      title = "Afgeleiden d(height)/d(year)",
      subtitle = "Vergelijking GAM vs GLM",
      x = "Jaar",
      y = "Afgeleide (trend per jaar)",
      color = "Modeltype",
      fill = "Modeltype"
    ) +
    theme(
      strip.text = element_text(size = 10),
      plot.title = element_text(size = 14, face = "bold")
    )
}

plot_derivatives(by_station_model_compared)



```

## AIC comparison
```{r}
by_station_model_compared %>%
  select(station, modeltype, adj.rsq, AIC) %>%
  ggplot(aes(station, AIC, color = modeltype)) +
  geom_point(shape = "|", size = 6) +
  coord_flip()
```

```{r}
plot_derivatives_glm <- function(by_station_model_compared) {

  # unnest derivatives én metadata
  df_plot <- by_station_model_compared %>% 
    select(station, modeltype, derivatives) %>%
    unnest(derivatives) %>%
    filter(modeltype == "glm")

  ggplot(df_plot, aes(x = x, y = derivative, color = station)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = station),
                alpha = 0.2, color = NA) +
    geom_line(size = 2) +
    # facet_wrap(~ station, scales = "free_y") +
    theme_bw() +
    labs(
      title = "Afgeleiden d(height)/d(year)",
      subtitle = "Vergelijking stations",
      x = "Jaar",
      y = "Afgeleide (trend per jaar)",
      color = "station"
    ) +
    theme(
      strip.text = element_text(size = 10),
      plot.title = element_text(size = 14, face = "bold")
    )
}

plot_derivatives_glm(by_station_model_compared)

```





## References

Keizer, Iris, Dewi Le Bars, Cees De Valk, André Jüling, Roderik Van De Wal, and Sybren Drijfhout. 2023. “The Acceleration of Sea-Level Rise along the Coast of the Netherlands Started in the 1960s.” Ocean Science 19 (4): 991–1007. https://doi.org/10.5194/os-19-991-2023.
