---
title: "GAM model estimation of Dutch Sea Level"
author: "Willem Stolte"
date: '`r Sys.Date()`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = FALSE
  )
```

## Introduction

Full documentation will follow

The Dutch Sea Level Monitor uses linear model variants to describe the sea level at the Dutch coastal tidal stations. In literature, other models are use. For example, GAM was used by Keizer et al., (2023) to detect changes in sea level rise. Although GAM is at the moment not included in the official product, in this document we test this method in order to visualize patterns of sea level (rise). 

This notebook provides the code and first results


```{r loadInformation}
# load necessary library
library(mgcv)    # for GAMs
library(tidyverse)
library(gratia)

epoch = 1970
datapath = "../../data/deltares/input/psmsl_gtsm_yr-latest.csv"

df <- read_delim(file.path(datapath), delim = ";", col_types = cols()) %>%
  # filter(station %in% params$station) %>%
  filter(year >= 1890)
```



## Define GAM 

The GAM model used here is using 4 components:

- smooth term
- nodal cosine term
- nodal sine term
- surge anomaly term

No further fine tuning of the GAM model has been done so far

```{r}

gam_model <- function(df, epoch = 1970){

  df <- df %>%
    mutate(
      nodal_cos = cos(2 * pi * (year - epoch) / 18.613),
      nodal_sin = sin(2 * pi * (year - epoch) / 18.613)
    )
  
  mgcv::gam(
    height ~ s(year, k = 5)   # smooth term on year
    +     nodal_cos
    +     nodal_sin
    + surge_anomaly,
    data = df,
    method = "REML"
  )
}

```

## Apply GAM model to sea level data for all stations

The GAM model was applied to all six main stations and the composite stations according to the code below. 

```{r }

selected_model = "gam"

by_station_model = df %>%
  group_by(station) %>%
  nest() %>%
  ungroup() %>%
  expand_grid(modeltype = selected_model) %>%
  #### parse function names based on selected model()
  mutate(modelfunctionname = paste(modeltype, "model", sep = "_")) %>%
  #### add functions for model calculation
  mutate(modelfunctions = map(modelfunctionname, get)) %>%
  #### add models based on data and functions
  mutate(model = pmap(
    list(
      data,
      modelfunctions
    ),
    \(.d, .f) .f(.d)
  )) %>%
  #### add output parameters and predictions
  mutate(
    glance = map(model, broom::glance),
    # rsq    = glance %>% map_dbl("r.squared"),
    # adj.rsq = glance %>% map_dbl("adj.r.squared"),
    # AIC    = glance %>% map_dbl("AIC"),
    tidy   = map(model, broom::tidy),
    augment = map(model, broom::augment)#,
    # equation = map(model, function(x) equatiomatic::extract_eq(x))
  ) %>%
  mutate(
    sm_predict = map(model, \(x) gratia::derivatives(x, select = "s(year)", type = "central", .name_repair = "universal")),
    
  ) %>%
  mutate(
    smoother = map(model, \(x) gratia::smooth_estimates(x, smooth = "s(year)")),
    Intercept = map(model, \(x) coef(x)["(Intercept)"])
  )


```

## Outcome of the model

```{r, results='asis'}

library(mgcv)
library(broom)
library(dplyr)

# Example GAM
# gam_model <- gam(y ~ s(x1) + s(x2) + z, data = mydata, method = "REML")

# Tidy coefficients
param_table <- broom::tidy(by_station_model$model[[1]])

# Extract goodness-of-fit from summary
gam_summary <- summary(by_station_model$model[[1]])

fit_stats <- tibble(
  deviance_explained = gam_summary$dev.expl,
  aic = AIC(by_station_model$model[[1]]),
  gcv_score = gam_summary$sp.criterion,  # Only if method = "GCV"
  n = gam_summary$n
)

# Combine
summary_table <- param_table %>%
  mutate(
    deviance_explained = fit_stats$deviance_explained,
    aic = fit_stats$aic,
    n = fit_stats$n
  )

summary_table

## still need to implement into tidy workflow below

summary_table <- by_station_model %>%
  select(station, glance, tidy) %>%
  unnest(c(glance, tidy)) %>%
  select(
    station,
    AIC,
    adj.r.squared,
    npar,
    p.value
  )

summary_table

```





### Skill assessment

```{r, eval = FALSE}
by_station_model %>%
  select(
    station, 
    glance
  ) %>%
  unnest(glance)
```

### Parameters

```{r}
by_station_model %>%
  select(
    station, 
    tidy
  ) %>%
  unnest(tidy)
```



## Sea level development in time

The comparison between observed and (GAM) modelled sea level for the 6 stations, and the composite stations is shown below. The variation in observations from 1950 to now is lower than in the period before. this is related to the wind-surge corrections derived from the GTSM model. Because of a lack of GTSM output for the years before 1950, for those years only a correction with the mean surge has been made. 

```{r, fig.width=8, fig.height=8, fig.cap="Relative sea level from the six Dutch main stations."}

by_station_model %>%
  unnest(data) %>%
  select(
    station,
    year,
    height,
    surge_anomaly # moet numeriek zijn!
  ) %>%
  ggplot(
    aes(
      x = year, 
      y = height - surge_anomaly
    )
  ) +
  geom_point(
    alpha = 0.6,
    aes(
      color = "observations"
    )) +
  geom_line(
    data = by_station_model %>%
      unnest(c(smoother, Intercept)),
    aes(
      x = year, 
      y = .estimate + Intercept,
      color = "gam"
    ),
    linewidth = 2,
    alpha = 0.6
  ) +
  # geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.3) +
  labs(x = "Year", y = "Estimated sea-level (mm)",
       title = "Evolution of sea-level from GAM model") +
  facet_wrap("station")

```

## Sea level rate development in time

```{r, fig.width=8, fig.height=7, fig.cap="Relative sea level change rate from the six Dutch main stations. The horizontal line is the current calculated sea level change rate as calculated by a broken-linear model in the Sea Level Monitor. "}
by_station_model %>%
  unnest(sm_predict) %>%
  ggplot(aes(x = year, y = .derivative)) +
  geom_line() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.3) +
  # geom_hline(yintercept = 3.1) +
  geom_linerange(xmin = 1993, xmax = 2025, y = 3.1, size = 1, color = "darkgreen") +
  annotate(x = 1990, y = 3.1 + 0.2, geom = "text", label = "ZSM2025", , color = "darkgreen") +
  # ylim(0,NA) +
  labs(x = "Year", y = "Estimated rate of sea-level rise (units per year)",
       title = "Smoother term of the evolution of the rate of sea-level rise from GAM model") +
  facet_wrap("station") +
  theme_minimal()


```


## References

Keizer, Iris, Dewi Le Bars, Cees De Valk, André Jüling, Roderik Van De Wal, and Sybren Drijfhout. 2023. “The Acceleration of Sea-Level Rise along the Coast of the Netherlands Started in the 1960s.” Ocean Science 19 (4): 991–1007. https://doi.org/10.5194/os-19-991-2023.
