---
title: "Virtual station analysis"
author: "Willem Stolte"
date: "`r Sys.Date()`"
output:
  github_document: null
  html_document:
    df_print: paged
    code_folding: hide
    outputdir: ../../docs
  pdf_document: default
always_allow_html: true
---


```{r setup,  echo=F, comment=F, include = F}

knitr::opts_chunk$set(
  echo = TRUE,
  comment=FALSE,
  message=FALSE,
  warning = FALSE
)

require(leaflet)
require(plotly)
require(tidyverse)
# require(reticulate)

list.files("")
config <- RcppTOML::parseToml("sealevelmonitor/_common/configuration.TOML")
config_flat <- list_flatten(config)
source("sealevelmonitor/_common/functions.R")
source("sealevelmonitor/_common/plotfunctions.R")
epoch = config$constants$epoch
mainstations_df <- readMainStationInfo(filepath = "../")
mainstations_locs <- readMainStationLocations(path = "../")
```

## Introduction

Sea level monitor uses a standardized method to determine the average sea level along the Dutch coast, and estimating the effect of nodal tide and storm surge. This method is averageing the sea level figures from the 6 main stations for all years from 1900 to present. This averages out any anomalies in individual stations that may have occurred due to changes in position or methods of the measuring station, and local effects such as morphological changes that may have influenced yearly mean sea level during periods.

## Overall methodology

In order to better estimate the sea level changes along the coast, we are testing to construct a "virtual" station, where the effect of anomalies that occur at separate stations are minimized. A simple way to accomplish this is to calculate median values (as opposed to mean values in the current SLM) of yearly sea levels for all stations. However, since the average sea levels between stations is different, inconsistencies can occur by taking median values of absolute sea level. Therefore, we first calculate the difference per year for each station, and then calculate the yearly difference for the "virtual" station by taking the median of all stations per year. From these differences, median sea level is reconstructed, using sea level at the start of the time series as a reference. A disadvantage is that the absolute vertical reference (e.g. to NAP) is lost during this exercise. It is therefore at this moment only an anlysis of the relative change of sea level.  


### Obtaining PSMSL "revised local reference" (RLR) water level data



```{r readPSMSLdata}
# get all Dutch PSMSL stations
ids <- c(mainstations_df$psmsl_id, 1551, 9, 236)
names <- c(mainstations_df$name, "Roompot Buiten", "Maassluis", "West-Terschelling")
slrData <- read_yearly_psmsl_csv(station_nr = ids, filepath = "../")
slrData2 <- slrData %>%
  select(
    year,
    rlr_height_mm,
    psmsl_id
  ) %>%
  left_join(
    tibble(id = ids, name = names),
    by = c(psmsl_id = "id")
  )

DT::datatable(slrData2)

```


### Data selection

Data are available for 9 stations, from approximately 1860 onward, depending on the location. In the Sea Level Monitor, only data from 1890 onward is used, because of a proper vertical reference before that time. We will here also select this period for further analysis. 


```{r rlr-plot, fig.cap="PSMSL obtained revised local reference water levels for all available Dutch stations. "}
slrData2 %>%
  ggplot(aes(year, rlr_height_mm)) +
  geom_line(aes(color = name)) +
  facet_wrap("name")

slrData2 <- slrData2 %>% filter(year >= 1890)
```

### Add surge 

GTSM surge values for each year is added to the dataframe. At the moment of this analysis, GTSM output for the 6 main stations was available. 

```{r getGTSM}

# Get GTSM data from local file
gtsm <- read_yearly_gtsm(filename = "../data/deltares/gtsm/gtsm_surge_annual_mean_main_stations.csv") |>
  mutate(year = year(ymd(t)))

```




```{r addGTSM, fig.width=8, fig.height=3, fig.cap="Surge anomalies (mm) for all available stations."}
  slrData3 <- slrData2 |>
      left_join(gtsm, by = c(name = "name", year = "year")) |>
      mutate(
        surge_anomaly = case_when(
          year >= 1950 ~ (1000 * surge - mean(1000 * surge, na.rm = T)), # meters to millimeters
          year < 1950 ~ 0
        )
      ) |>
  mutate(`height-surge anomaly` = rlr_height_mm - surge_anomaly) |>
      select(
        psmsl_id,
        year,
        rlr_height_mm,
        name,
        surge_anomaly,
        `height-surge anomaly`
      )

ggplot(
  slrData3 %>% 
    group_by(name) %>% 
    filter(
      sum(surge_anomaly)!=0,
      year > 1940
      ), 
  aes(year, surge_anomaly)) + 
  geom_line(aes(color = name))+
  theme(legend.position = "bottom")
```



### Calculate changes per year for all stations

For each station and year, the change of water level (corrected for surge) relative to the previous year is calculated. 


```{r plotDiff, fig.width=8, fig.height=3, fig.cap="Changes in sea level (surge corrected) per station and year (mm) for all available stations."}

# Met GTSM correctie
slrDiff2 <- slrData3 %>%
  filter(psmsl_id %in% mainstations_df$psmsl_id) %>%
  group_by(psmsl_id, name) %>%
  mutate(
    diff = `height-surge anomaly`-lag(`height-surge anomaly`, default=first(`height-surge anomaly`))
  ) %>% ungroup()

ggplot(slrDiff2, aes(year, diff)) +
  geom_line(aes(color = name)) +
   theme(legend.position = "bottom")
```


### calculate median slopes

For each year, the median change per year is calculated. By taking the median value, any outlier station values will not have an effect on the outcome.  

```{r makeMedianSlopes, fig.height=4, fig.width=8, fig.cap="Median change in surge corrected sea level for all stations (black line), per year. The change in surge corrected sea level for each station is added with colored lines. "}
slrMedian <- slrDiff2 %>%
  pivot_wider(id_cols = c(year), names_from = "name", values_from = "diff") %>%
  rowwise() %>%
  mutate(n = sum(!is.na(c_across(!year)))) %>%
  mutate(median = median(c_across(!year), na.rm = T)) %>%
  ungroup()  %>% 
  filter(
    year > 1890,
    n > 5
  )

ggplot(slrMedian, aes(year, median)) +
  geom_line(data = slrMedian %>%
           select(-n, -median) %>%
           pivot_longer(cols = -year, names_to = "station", values_to = "diff_mm"),
         aes(year, diff_mm, color = station), linewidth = 1, alpha = 0.5) +
  geom_line(linewidth = 1) +
   theme(legend.position = "bottom") +
  geom_vline(xintercept = 1932, linewidth = 1, color = "blue", alpha = 0.5) +
  annotate("text", x = 1935, y = 120, label = "1932")

cat("The mean value of the median slope over all stations and years is",
  round(mean(slrMedian$median), 2), 
  "+/-", 
  round(sd(slrMedian$median)),
  "mm per year")

```



### Results with linear model fit


```{r linear-with-tides}

slrMedian %>%
  mutate(`relative_median_height (mm)` = cumsum(median)) %>%
  ggplot(aes(year, `relative_median_height (mm)`)) +
  geom_line(aes()) +
  geom_smooth(method = "lm", formula = 
                y ~ I(x - 1970) #+ 
        # I(cos(2 * pi * (x - 1970)/(18.613))) + 
        # I(sin(2 * pi * (x - 1970)/(18.613)))
  )
  
```

### Results with broken linear model (breakpoint 1993)

```{r brokenlinear}

slrMedian %>%
  mutate(`relative_median_height (mm)` = cumsum(median)) %>%
  ggplot(aes(year, `relative_median_height (mm)`)) +
  geom_line(aes()) +
  geom_smooth(method = "lm", formula = 
                y ~ x +
                  I((x-1993)*(x > 1993)) #+
                # I(cos(2 * pi * (x - 1970)/(18.613))) +
                # I(sin(2 * pi * (x - 1970)/(18.613)))
  )

# lm(Value ~ Num*(Num >= 6.30) + Num*(Num < 6.30)
```

### Results with quadratic model (breakpoint 1960)

```{r}

slrMedian %>%
  mutate(`relative_median_height (mm)` = cumsum(median)) %>%
  ggplot(aes(year, `relative_median_height (mm)`)) +
  geom_line(aes()) +
  geom_smooth(method = "lm", formula = 
                y ~ x +
                  I((x-1960)*(x-1960)*(x > 1960)) #+
                # I(cos(2 * pi * (x)/(18.613))) +
                # I(sin(2 * pi * (x)/(18.613)))
  )

# lm(Value ~ Num*(Num >= 6.30) + Num*(Num < 6.30)
```

